### 2018-11-01 Sabanci_talk/Sabanci_talk.tex
More slides

\begin{frame}{Gravitational waves: d.o.f.'s}
$A_{\mu\nu} $ is the polarization tensor.
\begin{itemize}
 \uncover<2->{\item $\{ \bar{h}_{\mu\nu}, A_{\mu\nu} \}$: $4\times4$, symmetric \uncover<3->{$\hspace{1.5cm}\Rightarrow \f{4\times 5}{2}\ \ \ = 10$ d.o.f.}}
 \uncover<4->{\item Gauge condition: $\nabla_\mu \bar{h}^{\mu\nu} =  A_{\mu\nu} k^\nu= 0$ \uncover<5->{$\Rightarrow10-4 = \, 6$ d.o.f.}}
 \uncover<6->{\item Residual gauge freedom \hspace{2.25cm} \uncover<7->{$\Rightarrow 6-4 \ \ = \alert{2}$ \alert{d.o.f.}}}
 %\uncover<7->{\vspace{2mm}\begin{center}$ \bar{h}'_{\mu\nu} = \bar{h}_{\mu\nu} - \xi_{\mu,\nu} - \xi_{\nu,\mu} + \eta_{\mu\nu} \nabla_\alpha \xi^\alpha . $ \end{center}\vspace{2mm}}
\end{itemize}
\uncover<7->{\quad \ Consistent with $\pm 2$ helicities of a \alert{massless} spin 2 boson.\\}
\uncover<8->{$\Rightarrow$ \alert{2} POLARIZATIONS (\alert{transverse}): \uncover<10->{$\begin{array}{l}A_{xx}=-A_{yy} = h_\oplus  \\ A_{xy}=\ \ A_{yx} = h_\otimes  \\ \end{array}$}}
\\
\todo{Replace with Maggiore's discussion?}
{\begin{center} \includegraphics[height=2.79875cm]{figs/white_box.png}\end{center}}
\end{frame}

\begin{frame}{Gravitational waves: d.o.f.'s}
%\begin{center}
%\animategraphics[loop,controls,width=3cm]{12}{GWs_frame-}{0}{29} 
%\end{center}
$A_{\mu\nu} $ is the polarization tensor.
\begin{itemize}
{\item $\{ \bar{h}_{\mu\nu}, A_{\mu\nu} \}$: $4\times4$, symmetric {$\hspace{1.5cm}\Rightarrow \f{4\times 5}{2}\ \ \ = 10$ d.o.f.}}
 {\item Gauge condition: $\nabla_\mu \bar{h}^{\mu\nu} =  A_{\mu\nu} k^\nu= 0${ $\Rightarrow10-4 = \,6$ d.o.f.}}
 {\item Residual gauge freedom \hspace{2.25cm}{ $\Rightarrow 6-4 \ \ = \alert{2}$ \alert{d.o.f.}}}
 %{\vspace{2mm}\begin{center}$ \bar{h}'_{\mu\nu} = \bar{h}_{\mu\nu} - \xi_{\mu,\nu} - \xi_{\nu,\mu} + \eta_{\mu\nu} \nabla_\alpha \xi^\alpha . $ \end{center}\vspace{2mm}}
\end{itemize}
{\quad \ Consistent with $\pm 2$ helicities of a \alert{massless} spin 2 boson.\\}
{$\Rightarrow$ \alert{2} POLARIZATIONS (\alert{transverse}): {$\begin{array}{l}A_{xx}=-A_{yy} = h_\oplus  \\ A_{xy}=\ \ A_{yx} = h_\otimes  \\ \end{array}$}}
\\
\todo{Replace with Maggiore's discussion?}
{\begin{center} \animategraphics[loop,controls,height=2.4cm]{12}{GWs_plus-}{0}{20} \hspace{1cm} \animategraphics[loop,controls,height=2.4cm]{12}{GWs_cross-}{0}{20}\end{center}}
\end{frame}

@@

  Assumptions:
  \begin{itemize}
   \item Quasi-circular orbits: $e \ll 1$ and $\tfrac{|\dot{r}|}{r\Omega} <10^{-3} $ at $f=100\,$Hz.
   \item Point masses: tidal effects for $f \gtrsim 100\,$Hz
   \item Set $m_1 =m_2 = 1.4 M_\odot$, constraints: $1.2 M_\odot \lesssim m_\text{NS}\lesssim 1.6M_\odot$.
   \item Use Newtonian/post-Newtonian theory ($\le$3.5{\bf PN}).
   \item Terminate inspiral at Schwarzschild ISCO \todo{add ISCO curve?}
   \begin{footnotesize}
   \[f_\text{ISCO} = \f{c^3}{6^{3/2}\pi G M} \simeq 1571 \left(\f{2.8M_\odot}{M}\right)\text{Hz}\] 
   \end{footnotesize}
   \item Quadrupolar GWs only: higher modes suppressed by $\tfrac{m_1-m_2}{M}\tfrac{v^2}{c^2}$
   \item \todo{what about the 21 mode?}
   \item Neglect spins: spin-orbit $\tfrac{\dot{E}_\text{SO}}{\dot{E}} \sim \tfrac{v_s}{c}\tfrac{v}{c}\tfrac{R_\text{NS}}{r}$, spin-spin $\sim \tfrac{v^4}{c^4}$
  \end{itemize}

@@

\begin{frame}{Binary Neutron Star Inspirals}{Inspiral time}
Integrate $\dot{f} \sim f^{11/3} \quad \Longrightarrow \quad \int dt \sim \int f^{-11/3}df \quad\Rightarrow\quad t \sim \red{f^{-8/3}}$
\\
\vspace{1mm}
Fix integration constant $\tau = t_\text{coal} - t >0$. \\
\vspace{2mm}
Introduce \red{inspiral time}, i.e, time to BNS \bl{coal}escence
%
\begin{align}
 \red{\tau_\text{insp}(f)} &= \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}} \,{f^{-8/3}}\nn\\
 &\simeq \red{16.72\,\text{minutes}} \, \left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\,\red{\left(\f{10\,\text{Hz}}{f}\right)^{8/3}\quad(2)}\nn
\end{align}
Number of GW cycles to coalescence: $\int \tfrac{f}{\dot{f}}df\sim f^{-5/3}$ %$\int df {f}/\dot{f}$
\begin{align}
 \red{\mathcal{N}_\text{cyc}(f)} %&=\f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-5/3}-f_\text{max}^{-5/3}\right]\nn \\
 &\approx 1.605\times \red{10^4 \left(\f{10\,\text{Hz}}{f} \right)^{5/3}}\,\left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\quad \red{(3)}\nn
\end{align}

\begin{tikzpicture}[overlay,remember picture]
\node[draw, red,thick,minimum width=8.2cm,minimum height=1.3cm] (b) at (7.,3.35){};
\node[draw, red,thick,minimum width=10cm,minimum height=1.3cm] (b) at (6.2,1){};
\end{tikzpicture}

@@

>>>>

\begin{frame}{Gravitational waves: d.o.f.'s}
$A_{\mu\nu} $ is the polarization tensor.
\begin{itemize}
 \uncover<2->{\item $\{ \bar{h}_{\mu\nu}, A_{\mu\nu} \}$: $4\times4$, symmetric \uncover<3->{$\hspace{1.5cm}\Rightarrow \f{4\times 5}{2}\ \ \ = 10$ d.o.f.}}
 \uncover<4->{\item Gauge condition: $\nabla_\mu \bar{h}^{\mu\nu} =  A_{\mu\nu} k^\nu= 0$ \uncover<5->{$\Rightarrow10-4 = \, 6$ d.o.f.}}
 \uncover<6->{\item Residual gauge freedom \hspace{2.25cm} \uncover<7->{$\Rightarrow 6-4 \ \ = \alert{2}$ \alert{d.o.f.}}}
 %\uncover<7->{\vspace{2mm}\begin{center}$ \bar{h}'_{\mu\nu} = \bar{h}_{\mu\nu} - \xi_{\mu,\nu} - \xi_{\nu,\mu} + \eta_{\mu\nu} \nabla_\alpha \xi^\alpha . $ \end{center}\vspace{2mm}}
\end{itemize}
\uncover<7->{\quad \ Consistent with $\pm 2$ helicities of a \bl{massless} spin 2 boson.\\}
\uncover<8->{$\Rightarrow$ \alert{2} POLARIZATIONS (\alert{transverse}): \uncover<10->{$\begin{array}{l}A_{xx}=-A_{yy} = \red{h_+}  \\ A_{xy}=\ \ A_{yx} = \red{h_\times}  \\ \end{array}$}}
\\
\todo{Replace with Maggiore's discussion?}
{\begin{center} \includegraphics[height=2.79875cm]{figs/white_box.png}\end{center}}
\end{frame}

\begin{frame}{Gravitational waves: d.o.f.'s}
%\begin{center}
%\animategraphics[loop,controls,width=3cm]{12}{GWs_frame-}{0}{29} 
%\end{center}
$A_{\mu\nu} $ is the polarization tensor.
\begin{itemize}
{\item $\{ \bar{h}_{\mu\nu}, A_{\mu\nu} \}$: $4\times4$, symmetric {$\hspace{1.5cm}\Rightarrow \f{4\times 5}{2}\ \ \ = 10$ d.o.f.}}
 {\item Gauge condition: $\nabla_\mu \bar{h}^{\mu\nu} =  A_{\mu\nu} k^\nu= 0${ $\Rightarrow10-4 = \,6$ d.o.f.}}
 {\item Residual gauge freedom \hspace{2.25cm}{ $\Rightarrow 6-4 \ \ = \alert{2}$ \alert{d.o.f.}}}
 %{\vspace{2mm}\begin{center}$ \bar{h}'_{\mu\nu} = \bar{h}_{\mu\nu} - \xi_{\mu,\nu} - \xi_{\nu,\mu} + \eta_{\mu\nu} \nabla_\alpha \xi^\alpha . $ \end{center}\vspace{2mm}}
\end{itemize}
{\quad \ Consistent with $\pm 2$ helicities of a \bl{massless} spin 2 boson.\\}
{$\Rightarrow$ \alert{2} POLARIZATIONS (\alert{transverse}): {$\begin{array}{l}A_{xx}=-A_{yy} = \red{h_+}  \\ A_{xy}=\ \ A_{yx} = \red{h_\times}  \\ \end{array}$}}
\\
\todo{Replace with Maggiore's discussion?}
{\begin{center} \animategraphics[loop,controls,height=2.4cm]{12}{GWs_plus-}{0}{20} \hspace{1cm} \animategraphics[loop,controls,height=2.4cm]{12}{GWs_cross-}{0}{20}\end{center}}
\end{frame}

@@

  Assumptions:
  \begin{itemize}
   \item Quasi-circular orbits: $e \ll 1$ and $\tfrac{|\dot{r}|}{r\Omega} <10^{-3} $ at $f=100\,$Hz.
   \item Point masses: tidal effects for $f \gtrsim 100\,$Hz
   \item Set $m_1 =m_2 = 1.4 M_\odot$, constraints: $1.2 M_\odot \lesssim m_\text{NS}\lesssim 1.6M_\odot$.
   \item Use Newtonian/post-Newtonian theory ($\le$3.5{\bf PN}).
   \item Terminate inspiral at Schwarzschild ISCO \todo{add ISCO curve?}
   \begin{footnotesize}
   \[f_\text{ISCO} = \f{c^3}{6^{3/2}\pi G M} \simeq 1571 \left(\f{2.8M_\odot}{M}\right)\text{Hz}\] 
   \end{footnotesize}
   \item Quadrupolar GWs only: higher modes suppressed by $\tfrac{m_1-m_2}{M}\tfrac{v^2}{c^2}$
   \item Neglect spins: spin-orbit $\tfrac{\dot{E}_\text{SO}}{\dot{E}} \sim \tfrac{v_s}{c}\tfrac{v}{c}\tfrac{R_\text{NS}}{r}$, spin-spin $\sim \tfrac{v^4}{c^4}$
  \end{itemize}

@@

\begin{frame}{Binary Neutron Star Inspirals}{Inspiral time}
Integrate $\dot{f} \sim f^{11/3} \quad \Longrightarrow \quad \int dt \sim \int f^{-11/3}df \quad\Rightarrow\quad t \sim \red{f^{-8/3}}$
\\
\vspace{1mm}
Fix integration constant $\tau = t_\text{coal} - t >0$. \\
\vspace{2mm}
Introduce \red{inspiral time}, i.e, time to BNS \bl{coal}escence
%
\begin{align}
 \red{\tau_\text{insp}(f)} &= \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}} \,{f^{-8/3}}\nn\\
 &\simeq \red{16.72\,\text{minutes}} \, \left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\,\red{\left(\f{10\,\text{Hz}}{f}\right)^{8/3}\quad(2)}\nn
\end{align}
Number of GW cycles to coalescence: $\int \tfrac{f}{\dot{f}}df\sim f^{-5/3}$ %$\int df {f}/\dot{f}$
\begin{align}
 \red{\mathcal{N}_\text{cyc}(f)} %&=\f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-5/3}-f_\text{max}^{-5/3}\right]\nn \\
 &\approx 1.605\times \red{10^4 \left(\f{10\,\text{Hz}}{f} \right)^{5/3}}\,\left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\quad \red{(3)}\nn
\end{align}
%
\begin{tikzpicture}[overlay,remember picture]
\node[draw, red,thick,minimum width=8.2cm,minimum height=1.3cm] (b) at (7.,3.35){};
\node[draw, red,thick,minimum width=10cm,minimum height=1.3cm] (b) at (6.2,1){};
\end{tikzpicture}
%
\end{frame}

\begin{frame}{Gravitational waves from BNS}
TT gauge, multipole, slow-velocity expansions give \todo{add suppl. slide}
\[
 h_{ij}^\text{TT} = \tfrac{1}{D}\tfrac{2G}{c^4} \ddot{Q}_{il}^\text{TT}(t-z/c)
\]
%
 \begin{align}
 h_+(t) &= h_c(t)\, \left(\f{1+\cos^2\iota}{2}\right)\, \cos[\Phi(t)],\label{eq:hplus_TD}\\
 h_\times(t) & =h_c(t)\,\cos\iota \sin[\Phi(t)]\label{eq:hcross_TD},
\end{align}
%
%
where
%
\begin{align}
h_c(t) &=\f{4}{D}\left(\f{G M_c}{c^2}\right)^{5/3}\, \left(\f{\pi f(t)}{c}\right)^{2/3} \nn\\
\Phi(t) &=\int_{t_i}^t dt' \omega(t') + \text{PN corrections}
\end{align}

Frequecny increases as $f^{11/3}$\\
Amplitude increases as $f^{2/3} \Longrightarrow$ \red{Chirp!}

@@

\begin{frame}{Supplement 1: mode suppression}
\begin{itemize}
 \item For $q=1$ due to symmetry only (2,2), (3,2), (4,2), (4,4) are nonzero 
 \item For $q\ne 1$ all $\ell \ge 2$ modes exist, but are much smaller than $(2,2)$
\end{itemize}
